Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2018-06-12 LatticeOptics/latticeOpticsv5/LatticeOpticsV5.tex
Cleaned up figures

After the isolation stage, we use the sequence of optical elements shown in Fig.~\ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on the beam path) half-wave plate, and polarizer.
The Berek compensator is a $z$-cut quartz plate anti-reflection-coated at \unit{1064}{nm}.
By tilting the plate about its vertical axis by a few degrees, the extraordinary axis is mixed into the propagation of the beam, leading to  tilt-dependent birefringence that can be tuned from a zero-wave plate past a half-wave plate, represented by a retardation angle $\phi$.
Combined with the subsequent polarization optics, rotation of the Berek compensator can vary the transmitted power with a contrast of $\sim$1000:1.
Figure~\ref{fig:berek} shows the transmitted power as a function of retardation angle $\phi$ and fast-axis angle $\theta$ of the waveplates surrounding the compensator.
\begin{figure}
  \begin{center}
    \includegraphics{Figure4.pdf}
    \caption{\textbf{Berek compensator} transmission as a function of the Berek compensator's tilt angle $\phi$ and waveplate rotation angles $\theta$.}
    \label{fig:berek}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop}. This Berek compensator based feedback loop is used to vary the beam intensity at relatively high power. (a) Shows the small signal impulse response and (b-c) show the corresponding Bode plots.}\label{fig:berek_step_response}
  \end{center}
\end{figure}
A key point, as seen in Fig.~\ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator.
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in Fig.~\ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate.
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.
Since two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously.
To accomplish this, the low power loop is always manually railed to its high extreme when the high power loop is active, and the digital high power loop is disabled when the low power loop is active.

@@

\begin{figure}
  \begin{center}
    \includegraphics{fig/circuits.pdf}
    \caption{\textbf{Feedback circuits.} (a) The reverse-biased photodiode. (b) Externally referenced phase-locked loop. (c) Nonlinear loop filter. (d) RF power modulation using a mixer.}\label{fig:circuits}
  \end{center}
\end{figure}
The feedback circuit is shown in Fig.~\ref{fig:circuits}a. As seen in Fig.~\ref{fig:optical_layout}, a pair of low noise photodiodes measures the optical power of the beam sampled with a weakly reflective optic (Thorlabs BSF10-C).
The beam sampler is wedged in such a way that the two reflections diverge from one another, allowing us to illuminate both photodiodes.
One of the sampled beams is attenuated by an additional factor of $20\times$ by reflection off of an uncoated fused silica glass plate, such that the dynamic range of its photodiode covers $0-20$ W of the beam directed at the atoms, and controls the Berek compensator based, high-power feedback loop.
The other beam is not attenuated, save by the beam sampler, and its dynamic range covers $0-1$ W.
To prevent damage to the low power photodiode when the system is at its highest powers, a shutter (Stanford Research Systems SR475) blocks the light when the high power loop is in use.
The low power photodiode passes its signal $V_{\mathrm{pd}}$ to a loop filter circuit, which compares it to the control system's set-point, generated by digital to analog converter.
The loop filter's output controls the IF port of a microwave mixer (Minicircuits ZFM-2-S+), effectively modulating the RF power supplied by a low noise phase-locked loop (PLL).
The resulting RF signal is then amplified and sent to the AOM.
Let us now consider every part of this signal chain individually.

@@

\begin{figure}
  \begin{center}
    \includegraphics{fig/pll.pdf}
    \caption{\textbf{PLL performance} (a) The spectrum of the PLL, exhibiting phase noise of better than $-90$ dBc/Hz at $100$~kHz away from the carrier. (b) The RIN of the local oscillator based on the PLL. The noise level is seen to be at or below the noise floor of the measurement device.}\label{fig:pll}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{fig/cut_photodiode.jpg}
    \caption{\textbf{Photodiode photo} showing that the ``can'' of the photodiode has been cut, to prevent etalon-like behavior.}\label{fig:cut_pd}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics{fig/da_rin_scan.pdf}
    \caption{\textbf{Relative intensity noise of the digital to analog converter} at a fixed output voltage and while changing values. To measure the transient RIN, a $20$ ms, triangle wave with a DC offset and modulation depth of $15$\% of the DC offset was produced by the device. The triangle wave accounts for the series of spurs at low frequencies that fall with increasing frequencies. At frequencies above approximately $4$ kHz, the spectrum is dominated by noise due to transient ``glitches'' that emerge when the device changes its output state.}\label{fig:da_rin}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics{fig/low_power_bandwidths.pdf}
    \caption{\textbf{Loop bandwidth dependence on the set-point}. (a) shows the $3$ dB and $\pi$ phase shift points for the $X$ lattice. (b) shows the same for the $Y$ lattice.}\label{fig:bandwidth}
  \end{center}
\end{figure}

\begin{figure*}
  \begin{center}
    \includegraphics{fig/low_pass_filters.pdf}
    \caption{\textbf{Laser noise of the lattice system with and without additional low pass filters} (a) Closed loop laser made a fiber amplifier (different than the one in figure~\ref{fig:noises}. (b) Noise on the same system as (a), but with extra filters.}\label{fig:low_pass_noises}
  \end{center}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics{fig/AveragedMI.pdf}
    \caption{\textbf{Averaged images of Mott insulators}. a) Several Mott insulators were created over the course of approximately one hour, and the average image is shown\cite{Greif2016}. If the lattice fluctuated across more than a small fraction of a lattice site, the modulation contrast would be washed out. The absence of this effect means that the phase of the optical lattice is stable. b) shows an individual image of a Mott insulator.}\label{fig:averaged_mott}
  \end{center}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics{fig/pointing_figure.pdf}
    \caption{\textbf{Short term positional stability.} (a) A histogram of shot-to-shot step sizes along the transverse and longitudinal directions, with a Gaussian fit. (b) Correlations between the step size of subsequent shots. (c) Calibration data for the ``affine fit'' of the conversion from camera pixels to lattice dimentions. (d) Predicted location versus measured location along the transverse and longitudinal directions. (e,f) The measured position and predicted position of the atomic cloud, based on refitting the offset every $20$ experimental realizations.}\label{fig:stability}
  \end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\columnwidth]{fig/lattice_stability.pdf}
    \caption{\textbf{Long term stability over approximately nine months.} The lattice spacing $|\mathrm{v}_x|$ and $|\mathrm{v}_y|$ and the lattice angles $\theta_x$ and $\theta_y$ along the two lattice axes $x$ and $y$ are shown over the duration of almost 1 year. }\label{fig:long_term_stability}
  \end{center}
\end{figure}

>>>>

After the isolation stage, we use the sequence of optical elements shown in Fig.~\ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on the beam path) half-wave plate, and polarizer.
The Berek compensator is a $z$-cut quartz plate anti-reflection-coated at \unit{1064}{nm}.
By tilting the plate about its vertical axis by a few degrees, the extraordinary axis is mixed into the propagation of the beam, leading to  tilt-dependent birefringence that can be tuned from a zero-wave plate past a half-wave plate, represented by a retardation angle $\phi$.
Combined with the subsequent polarization optics, rotation of the Berek compensator can vary the transmitted power with a contrast of $\sim$1000:1.
Figure~\ref{fig:berek} shows the transmitted power as a function of retardation angle $\phi$ and fast-axis angle $\theta$ of the waveplates surrounding the compensator.
\begin{figure}
  \begin{center}
    \includegraphics{Figure4.pdf}
    \caption{\textbf{Berek compensator} transmission as a function of the Berek compensator's retardation angle $\phi$ and waveplate rotation angles $\theta$.}
    \label{fig:berek}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{Figure5.pdf}
    \caption{\textbf{Berek compensator based feedback loop}. This Berek compensator based feedback loop is used to vary the beam intensity at relatively high power. (a) Shows the small signal impulse response and (b-c) show the corresponding Bode plots.}\label{fig:berek_step_response}
  \end{center}
\end{figure}
A key point, as seen in Fig.~\ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator.
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in Fig.~\ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate.
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.
Since two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously.
To accomplish this, the low power loop is always manually railed to its high extreme when the high power loop is active, and the digital high power loop is disabled when the low power loop is active.

@@

\begin{figure}
  \begin{center}
    \includegraphics{Figure6.pdf}
    \caption{\textbf{Feedback circuits.} (a) The reverse-biased photodiode. (b) Externally referenced phase-locked loop. (c) Nonlinear loop filter. (d) RF power modulation using a mixer.}\label{fig:circuits}
  \end{center}
\end{figure}
The feedback circuit is shown in Fig.~\ref{fig:circuits}a. As seen in Fig.~\ref{fig:optical_layout}, a pair of low noise photodiodes measures the optical power of the beam sampled with a weakly reflective optic (Thorlabs BSF10-C).
The beam sampler is wedged in such a way that the two reflections diverge from one another, allowing us to illuminate both photodiodes.
One of the sampled beams is attenuated by an additional factor of $20\times$ by reflection off of an uncoated fused silica glass plate, such that the dynamic range of its photodiode covers $0-20$ W of the beam directed at the atoms, and controls the Berek compensator based, high-power feedback loop.
The other beam is not attenuated, save by the beam sampler, and its dynamic range covers $0-1$ W.
To prevent damage to the low power photodiode when the system is at its highest powers, a shutter (Stanford Research Systems SR475) blocks the light when the high power loop is in use.
The low power photodiode passes its signal $V_{\mathrm{pd}}$ to a loop filter circuit, which compares it to the control system's set-point, generated by digital to analog converter.
The loop filter's output controls the IF port of a microwave mixer (Minicircuits ZFM-2-S+), effectively modulating the RF power supplied by a low noise phase-locked loop (PLL).
The resulting RF signal is then amplified and sent to the AOM.
Let us now consider every part of this signal chain individually.

@@

\begin{figure}
  \begin{center}
    \includegraphics{Figure7.pdf}
    \caption{\textbf{PLL performance} (a) The spectrum of the PLL, exhibiting phase noise of better than $-90$ dBc/Hz at $100$~kHz away from the carrier. (b) The RIN of the local oscillator based on the PLL. The noise level is seen to be at or below the noise floor of the measurement device.}\label{fig:pll}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{Figure8.jpg}
    \caption{\textbf{Photodiode photo} showing that the ``can'' of the photodiode has been cut, to prevent etalon-like behavior.}\label{fig:cut_pd}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics{Figure9.pdf}
    \caption{\textbf{Relative intensity noise of the digital to analog converter} at a fixed output voltage and while changing values. To measure the transient RIN, a $20$ ms, triangle wave with a DC offset and modulation depth of $15$\% of the DC offset was produced by the device. The triangle wave accounts for the series of spurs at low frequencies that fall with increasing frequencies. At frequencies above approximately $4$ kHz, the spectrum is dominated by noise due to transient ``glitches'' that emerge when the device changes its output state.}\label{fig:da_rin}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics{Figure10.pdf}
    \caption{\textbf{Loop bandwidth dependence on the set-point}. (a) shows the $3$ dB and $\pi$ phase shift points for the $X$ lattice. (b) shows the same for the $Y$ lattice.}\label{fig:bandwidth}
  \end{center}
\end{figure}

\begin{figure*}
  \begin{center}
    \includegraphics{Figure11.pdf}
    \caption{\textbf{Laser noise of the lattice system with and without additional low pass filters} (a) Closed loop laser made a fiber amplifier (different than the one in figure~\ref{fig:noises}. (b) Noise on the same system as (a), but with extra filters.}\label{fig:low_pass_noises}
  \end{center}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics{Figure12.pdf}
    \caption{\textbf{Averaged images of Mott insulators}. a) Several Mott insulators were created over the course of approximately one hour, and the average image is shown\cite{Greif2016}. If the lattice fluctuated across more than a small fraction of a lattice site, the modulation contrast would be washed out. The absence of this effect means that the phase of the optical lattice is stable. b) shows an individual image of a Mott insulator.}\label{fig:averaged_mott}
  \end{center}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics{Figure13.pdf}
    \caption{\textbf{Short term positional stability.} (a) A histogram of shot-to-shot step sizes along the transverse and longitudinal directions, with a Gaussian fit. (b) Correlations between the step size of subsequent shots. (c) Calibration data for the ``affine fit'' of the conversion from camera pixels to lattice dimentions. (d) Predicted location versus measured location along the transverse and longitudinal directions. (e,f) The measured position and predicted position of the atomic cloud, based on refitting the offset every $20$ experimental realizations.}\label{fig:stability}
  \end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\columnwidth]{Figure14.pdf}
    \caption{\textbf{Long term stability over approximately nine months.} The lattice spacing $|\mathrm{v}_x|$ and $|\mathrm{v}_y|$ and the lattice angles $\theta_x$ and $\theta_y$ along the two lattice axes $x$ and $y$ are shown over the duration of almost 1 year. }\label{fig:long_term_stability}
  \end{center}
\end{figure}
